experDesign: stratifying samples into batches with minimal bias

### 2021-11-18 paper.Rmd
Remove figure 1, move text and thank editor and reviewers.

-   [*OSAT*](https://bioconductor.org/packages/OSAT/), at Bioconductor, first allocates the samples from each batch according to a variable; it then shuffles the samples from each batch in order to randomize the other variables [@yan2012].
    This algorithm relies on categorical variables and cannot use numerical variables (e.g., those that are age- or time-related) unless they are treated as categorical variables.

-   [*anticlust*](https://cran.r-project.org/package=anticlust), at CRAN, divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
    Conceptually it is similar to the clustering method k-means.

-   Recently, [*Omixer*](https://bioconductor.org/packages/Omixer/), a new package, has been made available at Bioconductor [@sinke2021].
    It tests whether the random assignments made by it are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.

@@

## OSAT

OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
Moreover, it works for both numeric and categorical variables but OSAT might return less rows than the input provided because they might have NA value.

@@

## anticlust

anticlust does not handle all types of variables, it only accepts numeric variables.

@@

## Omixer

There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
This precluded any comparisons of Omixer with other tools using the same settings.

@@

```
These methods results on this distributions:

```{r panel1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure A: distribution of  samples according to *OSAT*, Figure B according to *anticlust* on the variables used (1: Smoke, 2: Age, 3: Sex)."}
osat_plot <- osat_smoke / (osat_age + osat_sex)
osat_plot /anticlust + plot_layout(guides = 'collect') +
  plot_annotation(tag_levels = list(c("A.1", "A.2", "A.3", "B.2")),
                  title = "Existing solutions",
                  subtitle = "OSAT and anticlust solutions") &
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 8, hjust = 0, vjust = 0))
```

On Figure 1, we can see that the distributions is fairly similar between these two packages.

@@

```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with ***experDesign***.", warning=FALSE, message=FALSE}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations, name = "")
experDesign_index_spatial <- spatial(index = experDesign_index, 
                                     pheno = survey[, VoI], 
                                     iterations = iterations, 
                                     rows = LETTERS[1:8], 
                                     columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
# nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + 
                      experDesign_sex )) +
  plot_annotation(title = "experDesign") +
  plot_layout(guides = 'collect')
```

@@

We are grateful to Joe Moore for English-language assistance.

>>>>

-   [*OSAT*](https://bioconductor.org/packages/OSAT/), at Bioconductor, first allocates the samples from each batch according to a variable; it then shuffles the samples from each batch in order to randomize the other variables [@yan2012].
    This algorithm relies on categorical variables and cannot use numerical variables (e.g., those that are age- or time-related) unless they are treated as categorical variables.\
    OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
    Moreover, it works for both numeric and categorical variables but OSAT might return less rows than the input provided because they might have NA value.

-   [*anticlust*](https://cran.r-project.org/package=anticlust), at CRAN, divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
    Conceptually it is similar to the clustering method k-means.\
    anticlust does not handle all types of variables, it only accepts numeric variables.

-   Recently, [*Omixer*](https://bioconductor.org/packages/Omixer/), a new package, has been made available at Bioconductor [@sinke2021].
    It tests whether the random assignments made by it are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.\
    There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
    This precluded any comparisons of Omixer with other tools using the same settings.

@@

@@

@@

@@

```

@@

```{r experDesign, echo=FALSE, fig.cap="Example of distributions generated with ***experDesign***.", warning=FALSE, message=FALSE}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations, name = "")
experDesign_index_spatial <- spatial(index = experDesign_index, 
                                     pheno = survey[, VoI], 
                                     iterations = iterations, 
                                     rows = LETTERS[1:8], 
                                     columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
# nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + 
                      experDesign_sex )) +
  plot_annotation(title = "experDesign") +
  plot_layout(guides = 'collect')
```

On Figure 1 one can see that the distribution index generated by experDesign uses both numeric and categorical variables keeping also the samples with missing (NA) information.

@@

We are grateful to Joe Moore for English-language assistance and we would like to thank reviewers and the editor for taking the time and effort necessary to review the manuscript.
